---
title: "Group12-Project1"
author: "Group12"
date: "February 3, 2016"
output: html_document
---

## Improving Bike Storage Capacity

In effort to identify the bike stations that are candidates for improving bike storage capacity, we decided to investigate the difference between bike departure and bike arrivals for each station.

```{r}

library(plyr)
library(maps)

#read in citibike data
citibike <- read.csv("201512-citibike-tripdata.csv")

#attach citibike
attach(citibike)

#take a look at the head
head(citibike)

#convert station id to factor
start.station.id <- as.factor(start.station.id)
end.station.id <- as.factor(end.station.id)

#
numStartStation <- count(start.station.id)
colnames(numStartStation) <- c("station_id", "start_freq")
numEndStation <- count(end.station.id)
colnames(numEndStation) <- c("station_id", "end_freq")

stationCount <- merge(numStartStation, numEndStation, sort = FALSE, by = "station_id")

stationCount$dep_minus_arr <- stationCount$start_freq - stationCount$end_freq
stationCount <- arrange(stationCount, stationCount$dep_minus_arr)
```

```{r}
#build a table for bike station id to longitude/latitude
uniqueStation <- citibike[!duplicated(citibike$start.station.id),c("start.station.id", "start.station.latitude", "start.station.longitude")]
colnames(uniqueStation) <- c("station_id", "latitude", "longitude")

#number of stations being used in December 2015
nrow(uniqueStation)
```

```{r}
#get top 10 stations that have more arrivals than departures
mostArrivalStations <- head(stationCount, 10)
mostArrivalStations
#get top 10 stations that have more departures than arrivals
mostDepartureStations <- tail(stationCount, 10)
mostDepartureStations
longitude = c()
latitude = c()
for (i in 1:length(mostArrivalStations$station_id)){
  longitude[i] = citibike$end.station.longitude[match(x = mostArrivalStations$station_id[i], table = citibike$end.station.id)]
  latitude[i] = citibike$end.station.latitude[match(x = mostArrivalStations$station_id[i], table = citibike$end.station.id)]
}

mostArrivalStations$longitude = longitude
mostArrivalStations$latitude = latitude

longitude = c()
latitude = c()
for (i in 1:length(mostDepartureStations$station_id)){
  longitude[i] = citibike$start.station.longitude[match(x = mostDepartureStations$station_id[i], table = citibike$start.station.id)]
  latitude[i] = citibike$start.station.latitude[match(x = mostDepartureStations$station_id[i], table = citibike$start.station.id)]
}

mostDepartureStations$longitude = longitude
mostDepartureStations$latitude = latitude

#plot(numStartStation$freq ~ numStartStation$x)
```

To help visualize our findings, we will plot the stations with the most arrivals relative to departures as red points and the most departures relative to arrivals as blue points. Points with higher alpha indicate a larger difference between departure and arrivals.

```{r}
library(ggplot2)
library(ggmap)
nymap <- get_map(location = "New York", zoom = 12, color = 'bw')

#map the stations with the most arrivals relative to departures
maxArrWeight <- max(-mostArrivalStations$dep_minus_arr)
mostArrAlpha <- abs(mostArrivalStations$dep_minus_arr)/maxArrWeight
maxDepWeight <- max(mostDepartureStations$dep_minus_arr)
mostDepAlpha <- mostDepartureStations$dep_minus_arr/maxDepWeight

mapPoints <- ggmap(nymap) + geom_point(aes(x = longitude, y = latitude), size = 6*mostArrAlpha, colour = 'red', data = mostArrivalStations) + geom_point(aes(x = longitude, y = latitude), size=6*mostDepAlpha, colour = 'blue', data = mostDepartureStations)
mapPoints 
```

Naively, we can say that the red points are the best candidates for an increase in capacity since they have the most surplus in arrivals.

## Reducing Bike Maintenance Bills

Next, to address the increase in bike maintenance bills, we will first visualize the bike usage by bike ID. First we take a look at the total number of trips involving each bike.

```{r}
attach(citibike)
bikeCount <- count(bikeid)

#Number of unique bikes used over the month of December 2015
print(nrow(bikeCount))

#Analyze frequency 
summary(bikeCount$freq)
boxplot(bikeCount$freq, outline = FALSE, horizontal = TRUE, main = "Frequency of bike usage in December 2015", xlab = "Number of trips")
```
From the summary and graph above, we see that out of the `r nrow(bikeCount)` bikes, 

However, not all trips have the same duration. So we chose to analyze each bike's total trip duration throughout the month of December 2015.
```{r}
attach(citibike)
bikeUsageById <- tapply(tripduration, start.station.id, sum, na.rm = TRUE)
bikeUsageById <- bikeUsageById/3600
summary(bikeUsageById)
boxplot(bikeUsageById, horizontal = TRUE, main = "Frequency of total bike duration in December 2015", xlab = "Accumulated Bike Duration (hours)")
```

From the graph above, it is easy to see that the frequency distribution is skewed to the right - meaning there are a few bikes with a much higher accumulated bike durations than the rest. Just by observing the box plot we can reasonably infer that some bikes are being used much more than other bikes. In fact the standard deviation of the total accumulated duration per bike is `r sd(bikeUsageById)`, which is significantly higher than 0, which is what we expect if all bikes truly had equal usage.

## Youth and Tourist Promotion

First we identify the bike stations that have the least youthful attendance.

```{r}
citibike$isYoung <- ifelse(test = is.na(citibike$birth.year), yes = FALSE, no = ((2016 - citibike$birth.year) <= 25))
  
stationYouthCount <- aggregate(x = citibike$isYoung, by = list(citibike$start.station.id), FUN = sum)
colnames(stationYouthCount) <- c("station_id", "count")
stationYouthCount <- stationYouthCount[order(stationYouthCount$count),]

longitude = c()
latitude = c()
for (i in 1:length(stationYouthCount$station_id)){
  longitude[i] = citibike$start.station.longitude[match(x = stationYouthCount$station_id[i], table = citibike$start.station.id)]
  latitude[i] = citibike$start.station.latitude[match(x = stationYouthCount$station_id[i], table = citibike$start.station.id)]
}
stationYouthCount$longitude = longitude
stationYouthCount$latitude = latitude

#get the top 20 bike stations with the least amount of youth
leastYouthfulStations <- head(stationYouthCount, n = 20)

leastYouthMap <- get_map(location = "New York", zoom = 12, color = 'bw')

#map the locations of the least youthful stations

leastYouthfulPoints <- ggmap(leastYouthMap) + geom_point(aes(x = longitude, y = latitude),  colour = 'dark green', data = leastYouthfulStations)
leastYouthfulPoints
```

Next we identify the stations less used by tourists

```{r}  
stationTourist <- aggregate(x = citibike$usertype == "Customer", by = list(citibike$start.station.id), FUN = sum)
colnames(stationTourist) <- c("station_id", "count")
stationTourist <- stationTourist[order(stationTourist$count),]

longitude = c()
latitude = c()
for (i in 1:length(stationTourist$station_id)){
  longitude[i] = citibike$start.station.longitude[match(x = stationTourist$station_id[i], table = citibike$start.station.id)]
  latitude[i] = citibike$start.station.latitude[match(x = stationTourist$station_id[i], table = citibike$start.station.id)]
}
stationTourist$longitude = longitude
stationTourist$latitude = latitude

#get the top 20 bike stations with least amount of tourists
stationLeastTuorist <- head(stationTourist, n = 20)

leastTouristMap <- get_map(location = "New York", zoom = 12, color = 'bw')

#map the locations of the least youthful stations

leastTouristPoints <- ggmap(leastTouristMap) + geom_point(aes(x = longitude, y = latitude),  colour = 'dark blue', data = stationLeastTuorist)
leastTouristPoints
```


### Analysis by time of day
```{r}
library(lubridate)
bikeStartTime <- as.character(citibike$starttime)
bikeStartTime <- strptime(bikeStartTime, "%m/%d/%Y %H:%M:%S")
bikeStartByHour <- hour(bikeStartTime)
bikeByHourCnt <- count(bikeStartByHour)
colnames(bikeByHourCnt) <- c("Time of Day", "Number of Bike Rides")
barplot(bikeByHourCnt[,"Number of Bike Rides"], names.arg = bikeByHourCnt[,"Hour in Day"], main = "Bike Rides by Time of Day", xlab = "Time of Day", ylab = "Number of Bike Rides", col = "light blue")
```

The result interesting because we see a bi-modal distribution that peaks during 8 a.m. - 9 a.m. in the morning 5 p.m. - 6 p.m. in the evening. This makes it pretty apparent that bikes are used most often to commute to and from work.

But what about bike usage by time during weekends? We decided to perform the same analysis above, but this time first partitioning the data set into weekends and weedays.

```{r}
isWeekend <- bikeStartTime$wday >= 6
bikeStartWeekend <- bikeStartByHour[c(isWeekend)]
bikeStartWeekday <- bikeStartByHour[c(!isWeekend)]

d <- count(bikeStartWeekend)
bikeByHourWeekday <- count(bikeStartWeekday)

colnames(bikeByHourWeekend) <- c("Time of Day", "Number of Bike Rides")
colnames(bikeByHourWeekday) <- c("Time of Day", "Number of Bike Rides")

barplot(bikeByHourWeekend[,"Number of Bike Rides"], names.arg = bikeByHourWeekend[,"Time of Day"], main = "Weekend: Bike Rides by Time of Day", xlab = "Time of Day", ylab = "Number of Bike Rides", col = "light green")

barplot(bikeByHourWeekday[,"Number of Bike Rides"], names.arg = bikeByHourWeekday[,"Time of Day"], main = "Weekday: Bike Rides by Time of Day", xlab = "Time of Day", ylab = "Number of Bike Rides", col = "violet")

```

The data seems to substantiate our claim that most bike rides are used to commute to and from work since we still observe the bi-modal distribution on the bike trip frequency graph over weekdays. However on weekends, 



